---
title: "R Coding Exercise"
author: 'William Hyltin'
date: today
output: html_document
---
This project is intended to serve as R coding practice, both for familiarity with R language and for some practice documenting code. We will be using the gapminder health and income dataset.

Loading necessary packages for the project, I like to use `pacman::p_load()` because it checks for installation of packages before loading the library and automatically installs ones that I don't have, as well it allows me to load multiple in just one line. The only major downside is that it does require the installation of the `pacman` package to use. \

-   `dslabs` in order to get the gapminder dataset.\
-   `tidyverse` for its convenience in data cleaning and visualization.  

```{r setup}
pacman::p_load(dslabs, tidyverse, readxl)
# I have the library versions commented out here for anyone needing to 
# replicate this code that doesnt have the pacman package installed.

#library(dslabs)
#library(tidyverse)
#library(readxl)
```

Now we can take a look at the gapminder dataset. We start by using the `str()` function to get the Structure of the data. Then use the `summary()` function to get a quick summary of each of the variables in the dataset. Finally we use the `class()` function to confirm what type of object the gapminder dataset is.  
```{r}
#help(gapminder) #commented out for the sake of rendering later
print('-----Data Structure-----') #these are just to make the outputs a little more readable.
str(gapminder)

print('-----Data Summary-----')
summary(gapminder)

print('-----Object Type-----')
class(gapminder)
```

Now we can filter to the continent variable to just Africa. To do this I use the `dplyr` function `filter()` along with the pipe `%>%`. I primarily do this for readability, although subsetting with base R syntax would be just as easy to do. After that I use `str()` and `summary()` again to get the data structure and summaries.  
```{r}
africadata <- gapminder %>% filter(continent == 'Africa')

print('-----Data Structure-----')
str(africadata)

print('-----Data Summary-----')
summary(africadata)
```

Now we will create two new objects, `im.le` and `pop.le` to isolate variables of interest.  

* `im.le` will consist of the variables `infant_mortality` and `life_expectancy`  
* `pop.le` will consist of the variables `population` and `life_expectancy`  

We will accomplish this by using again the pipe operator `%>%` and the `select()` function.  
```{r}
im.le <- africadata %>% select(infant_mortality, life_expectancy) 
  # select() allows us to choose relevant columns for our new objects
pop.le <- africadata %>% select(population, life_expectancy)

print('-----Data Structure-----')
print('---Population x Life Expectancy---')
str(im.le)
print('---Infant Mortality x Life Expectancy---')
str(pop.le)

print('-----Data Summary-----')
print('---Population x Life Expectancy---')
summary(im.le)
print('---Infant Mortality x Life Expectancy---')
summary(pop.le)
```

```{r}
im.le %>% 
  ggplot(aes(x=infant_mortality, y=life_expectancy)) +
  geom_point() +
  labs(x='Infant Mortality', y='Life Expectancy', 
       title='Examination of Life Expectancy in African Countries', 
       subtitle='Relationship of Infant Mortality and Life Expectancy')

pop.le %>% 
  ggplot(aes(x=population, y=life_expectancy)) +
  coord_trans(x='log2') +
  geom_point() +
  labs(x='Population', y='Life Expectancy', 
       title='Examination of Life Expectancy in African Countries', 
       subtitle='Relationship of Population and Life Expectancy')
```

The charts imply a negative correlation between infant mortality and life expectancy, as well as a positive correlation between population and life expectancy. However the "streaks" of points that we are seeing are likely due to the `year` variable from our original dataset. What we are seeing is the year over year change in each country's life expectancy and infant mortality or population. We can isolate to one year in particular to avoid this. First we should determine which years have missing data for `infant_mortality`.
```{r}
africadata %>% 
  group_by(year) %>% #group_by function will allow us to easily identify the year
  summarize(
    missing_im = sum(is.na(infant_mortality)) #takes advantage of sum function and logical values since TRUE==1
  )
```

As we can see, there are missing values for infant mortality all the way up to 1981, and again in 2016. So we will just need to choose a year after 1981, but not 2016.

We will isolate to the year 2000 by using the dplyr `filter()` function again.
```{r}
im.le2000 <- africadata %>% 
  filter(year == 2000) %>% 
  select(infant_mortality, life_expectancy)

pop.le2000 <- africadata %>% 
  filter(year == 2000) %>% 
  select(population, life_expectancy)

print('-----Data Structure-----')
print('---Population x Life Expectancy---')
str(im.le2000)
print('---Infant Mortality x Life Expectancy---')
str(pop.le2000)

print('-----Data Summary-----')
print('---Population x Life Expectancy---')
summary(im.le2000)
print('---Infant Mortality x Life Expectancy---')
summary(pop.le2000)
```

Repeating the process for our plots with the new data:
```{r}
im.le2000 %>% 
  ggplot(aes(x=infant_mortality, y=life_expectancy)) +
  geom_point() +
  labs(x='Infant Mortality', y='Life Expectancy', 
       title='Examination of Life Expectancy in African Countries', 
       subtitle='Relationship of Infant Mortality and Life Expectancy During the Year 2000')

pop.le2000 %>% 
  ggplot(aes(x=population, y=life_expectancy)) +
  coord_trans(x='log2') +
  geom_point() +
  labs(x='Population', y='Life Expectancy', 
       title='Examination of Life Expectancy in African Countries', 
       subtitle='Relationship of Population and Life Expectancy During the Year 2000')
```

We see there is still likely a relationship between infant mortality and life expectancy, but the relationship between population and life expectancy is less apparent, if it is still there at all. To get more conclusive results, we can fit a linear model between the variables using the `lm()` function.

```{r}
fit1 <- lm(life_expectancy~., data = im.le2000)
#fits life expectancy as a function of infant mortality. 
#Infant mortality is not explicitly listed since it is the only other variable in the object.

fit2 <- lm(life_expectancy~., data = pop.le2000) 
#fits life expectancy as a function of population. 
#Population is not explicitly listed since it is the only other variable in the object.

print('-----Life Expectancy as a function of Infant Mortality-----')
summary(fit1) #summary of each linear model fit to get results.

print('-----Life Expectancy as a function of Population-----')
summary(fit2)

```

For the fit of Life Expectancy as a function of Infant Mortality, we see a p-value of 2.826e-08. Against an alpha-level of 0.05, we would conclude that there is a significant linear relationship between Infant Mortality and Life Expectancy.  

For the fit of Life Expectancy as a function of Population, we see a p-value of 0.6159. Against an alpha-level of 0.05, we would conclude that there is not a significant linear relationship between Population and Life Expectancy.  

## The following section is contributed by ZANE CHUMLEY.

### Pick a dataset

```{r}
# It's an election year, so let's look at the polls from the year Trump eventually won.
ZaneA03 <- results_us_election_2016
```

### Explore the dataset

```{r}
# look at the data's type
class(ZaneA03)

# look at the data's structure
str(ZaneA03)

# look at a summary of the data
summary(ZaneA03)
```

It is worth noting that the dataset is significantly less detailed than described in https://cran.r-project.org/web/packages/dslabs/dslabs.pdf.  While there are only 5 columns in the dataset, the description indicated many more columns would be provided:

+ state. State in which poll was taken. ‘U.S‘ is for national polls.
+ startdate. Poll’s start date.
+ enddate. Poll’s end date.
+ pollster. Pollster conducting the poll.
+ grade. Grade assigned by fivethirtyeight to pollster.
+ samplesize. Sample size.
+ population. Type of population being polled.
+ rawpoll_clinton. Percentage for Hillary Clinton.
+ rawpoll_trump. Percentage for Donald Trump
+ rawpoll_johnson. Percentage for Gary Johnson
+ rawpoll_mcmullin. Percentage for Evan McMullin.
+ adjpoll_clinton. Fivethirtyeight adjusted percentage for Hillary Clinton.
+ ajdpoll_trump. Fivethirtyeight adjusted percentage for Donald Trump
+ adjpoll_johnson. Fivethirtyeight adjusted percentage for Gary Johnson
+ adjpoll_mcmullin. Fivethirtyeight adjusted percentage for Evan McMullin

### Do any processing/cleaning you want to do

From the exploration above it does not appear there are any NA values in the data.  Let's check to be sure.

```{r}
# Any NA values in the state column?
ZaneA03.state.NAs <- ZaneA03[ZaneA03$state=="NA",]
str(ZaneA03.state.NAs)

# Any NA values in the electoral_votes column?
ZaneA03.electorals.NAs <- ZaneA03[ZaneA03$electoral_votes=="NA",]
str(ZaneA03.electorals.NAs)

# Any NA values in the clinton column?
ZaneA03.clinton.NAs <- ZaneA03[ZaneA03$clinton=="NA",]
str(ZaneA03.clinton.NAs)

# Any NA values in the trump column?
ZaneA03.trump.NAs <- ZaneA03[ZaneA03$trump=="NA",]
str(ZaneA03.trump.NAs)

# Any NA values in the others column?
ZaneA03.others.NAs <- ZaneA03[ZaneA03$others=="NA",]
str(ZaneA03.others.NAs)

```

There are no objects in any of the five (5) datasets housing NA values.  Therefore, no cleaning is warranted.

But are there any outliers?

### Make a few exploratory figures. 

```{r}
# Let's use boxplots to see if there are any outliers in the four (4) columns containing numerical data
boxplot(ZaneA03$electoral_votes
        , main="Boxplot of Electorcal Votes"
        , xlab=""
        , horizontal=TRUE
        )
boxplot(ZaneA03$clinton
        , main="Boxplot of Clinton poll readings"
        , xlab=""
        , horizontal=TRUE
        )
boxplot(ZaneA03$trump
        , main="Boxplot of Trump poll readings"
        , xlab=""
        , horizontal=TRUE
        )
boxplot(ZaneA03$others
        , main="Boxplot of pool readings for other candidates"
        , xlab=""
        , horizontal=TRUE
        )
```

### Optionally, also some tables.

```{r}
# Let's display all the rows containing an outlier revealed by the boxplots above.

# We'll sorting the rows by the values in each column into new datasets.
# The sorting will be largest values first.
# Then we will display the top and/or the bottom of the dataset corresponding to the upper and lower outliers, respectively.
ZaneA03.electorals.sorted <- ZaneA03[order(-ZaneA03$electoral_votes),]
head(ZaneA03.electorals.sorted
     , n=3
     )
ZaneA03.clinton.sorted <- ZaneA03[order(-ZaneA03$clinton),]
head(ZaneA03.clinton.sorted
     , n=1
     )
ZaneA03.trump.sorted <- ZaneA03[order(-ZaneA03$trump),]
tail(ZaneA03.trump.sorted
     , n=1
     )
ZaneA03.others.sorted <- ZaneA03[order(-ZaneA03$others),]
head(ZaneA03.others.sorted
     , n=5
     )
```

### Run some simple statistical model(s). Your choice.

How successful were the polls in predicting which candidate ultimately carried the state in the election?  Well, we'll need another dataset ... the results from the voting.

```{r}
# Load actual votes from 2016 
ZaneA03.votedata <- read_xlsx("1976-2020-president.xlsx"
                              , sheet="2016Flat"
                              , col_names = TRUE
                              )
# look at the data's type
class(ZaneA03.votedata)

# look at the data's structure
str(ZaneA03.votedata)

# look at a summary of the data
summary(ZaneA03.votedata)

```

```{r}
# Are we fortunate enough that the dataset of polls and the dataset of votes are in the same order by state?   One way to check is through visual inspection.

ZaneA03.state.sorted <- ZaneA03[order(ZaneA03$state),]
ZaneA03.sortcheck <- rbind(ZaneA03.state.sorted$state
                           , ZaneA03.votedata$State
                            )
head(ZaneA03.sortcheck
     , n=51
     )

```

The visual inspection reveals the data is aligned by state.

```{r}
# The visual inspection reveals the data is aligned by state. 
# So, polls and votes, by the power invested in me by R,
# I pronounce your merged!

# Build the list of datasets to merge

# ZaneA03.datasetlist <- c("ZaneA03.state.sorted"
#                         , "ZaneA03.votedata"
#                         )
# class(ZaneA03.datasetlist)
```

```{r}
# Go forth and merge!

# Rest of this section commented out so you can see everything above in the render
# ZaneA03.PollsAndVotes <- rbind(ZaneA03.state.sorted
#                               , ZaneA03.votedata)
```

```{r}
# 
#ZaneA03.PollsAndVotes <- ZaneA03.state.sorted + ZaneA03.votedata

```
